---
title: "Control access to CerebroNexus with a login page"
author:
  - "Roman Hillje"
  - "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
toc-title: "Contents"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Control access to CerebroNexus with a login page}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<style>
.main-container {
  width: auto !important;
  max-width: none !important;
  margin: 0;
  padding: 0 clamp(1rem, 4vw, 4rem);
}
pre {
  overflow-x: auto;
}
img {
  display: block;
  width: 100%;
  height: auto;
  margin: 1.5rem auto;
}
</style>

# Overview

`createShinyApp()` creates a public Viewer unless you pass `auth`. Auth lite
expects two existing inputs:

1. an encrypted `shinymanager` database; and
2. an environment variable containing that database's passphrase.

The App receives the encrypted database. Passwords and the passphrase stay
outside the App. HTTPS is still required.

![Auth lite checks the descriptor, environment, and database before publication
and again when the Viewer starts.](img/auth-lite-validation.svg)

# One-minute setup

**Run this as:** the person preparing the Viewer release.

**Run it on:** one trusted machine with R, CerebroNexus, and the source CRB files.

**What happens:** you create one encrypted credentials database, one external
secret file, and one login-protected App directory.

**Verify:** the final `createShinyApp()` call must succeed before anything is
deployed; then test one valid and one invalid login locally.

Install the optional provider:

```{r eval=FALSE}
# R console — run this once to install and load the authentication tools.
install.packages("shinymanager")
library(CerebroNexus)
```

Generate a database passphrase on a trusted machine:

```bash
# Terminal — run this command and save the result privately.
openssl rand -base64 32
```

Prepare the private and App directories first. Run this once as the server
operator; it makes the release account the owner so the subsequent R commands
can write under `/srv` without elevated privileges:

```bash
# Terminal — run this once on the trusted build or deployment machine.
sudo install -d -m 0700 -o "$USER" -g "$(id -g)" /srv/cerebro/private
sudo install -d -m 0755 -o "$USER" -g "$(id -g)" /srv/cerebro/apps
```

Create the environment file outside the App directory:

```{r eval=FALSE}
# R console — run this to create the external env file with private permissions.
secret_file <- "/srv/cerebro/private/viewer-auth.env"
stopifnot(!file.exists(secret_file))

old_umask <- Sys.umask("077")
tryCatch(
  {
    stopifnot(file.create(secret_file))
    writeLines(
      "CEREBRO_AUTH_PASSPHRASE=replace-with-the-random-value",
      secret_file
    )
  },
  finally = Sys.umask(old_umask)
)
Sys.chmod(secret_file, "0600")
```

Create the encrypted database:

```{r eval=FALSE}
# R console — run this to create the encrypted user database.
readRenviron(secret_file)
accounts <- data.frame(
  user = c("alice", "bob"),
  password = c("alice-login-password", "bob-login-password"),
  admin = c(TRUE, FALSE),
  stringsAsFactors = FALSE
)

shinymanager::create_db(
  credentials_data = accounts,
  sqlite_path = "/srv/cerebro/private/credentials.sqlite",
  passphrase = Sys.getenv("CEREBRO_AUTH_PASSPHRASE")
)
rm(accounts)
```

Build the protected Viewer:

```{r eval=FALSE}
# R console — run this to build the login-protected Viewer.
createShinyApp(
  cerebro_data = c("My dataset" = "output/cerebro_my_dataset.crb"),
  result_dir = "/srv/cerebro/apps/my_app",
  port = 3838,
  auth = list(
    credentials = "/srv/cerebro/private/credentials.sqlite",
    passphrase_env = "CEREBRO_AUTH_PASSPHRASE",
    timeout_minutes = 15
  )
)
```

Start the Viewer and clear the process variable when finished:

```{r eval=FALSE}
# R console — run this to start the generated App locally.
readRenviron("/srv/cerebro/private/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/my_app")
Sys.unsetenv("CEREBRO_AUTH_PASSPHRASE")
```

# What goes where

| Item | Correct location | In the App? |
|---|---|---:|
| Source `credentials.sqlite` | Private build area | No; a copy is added |
| `viewer-auth.env` | Secret store or protected path | **Never** |
| Generated App | Deployment directory | Yes |
| Bundled encrypted database | `private-data/auth/` | Yes |
| Plaintext account table | Trusted R memory only | **Never** |

The runtime account needs only read access to the bundled database. At startup,
the Viewer copies it into a private temporary directory for `shinymanager`
login logs, failed-password counters, and password changes. That runtime state
lasts only for the life of the Viewer process. Update durable accounts or
passwords in the private source database, then rebuild and redeploy the App.

![The App and external passphrase travel separately and meet only in the trusted
runtime process.](img/auth-deployment-boundary.svg)

# Validation rules

The build stops before publication unless all checks pass:

The names(auth) must be exactly "credentials", "passphrase_env", and optionally "timeout_minutes".

- `shinymanager >= 1.1.0` is installed;
- `auth` has only supported fields;
- the database is a readable regular SQLite file;
- the named environment variable exists and is at least 16 characters;
- `timeout_minutes` is a whole number from 1 to 1440; and
- the database decrypts and has the expected tables and hashed passwords.

A wrong passphrase never produces a half-configured App.

# Run locally

**Run this as:** the account that will start the App.

**Run it on:** the machine holding the generated App and its matching external
env file.

**What happens:** `readRenviron()` places the passphrase in this R process;
`runApp()` reads it to unlock the bundled database and start the login page.

**Verify:** open <http://localhost:3838>, confirm a valid login succeeds and an
invalid login fails, then stop the R process with Ctrl+C.

Every new R process must load the external secret:

```{r eval=FALSE}
# R console — run this in every local App process.
options("shinymanager.pwd_failure_limit" = 10L)
readRenviron("/srv/cerebro/private/viewer-auth.env")
shiny::runApp("/srv/cerebro/apps/my_app")
```

Check one valid and one invalid login in a private browser window. Repeat the
check on the real HTTPS URL.

# Deploy with Shiny Server and systemd

**Run this as:** the server operator; use elevated privileges only for file and
service administration.

**Run it on:** the Shiny Server host, not on the build workstation.

**What happens:** the App is installed under the Shiny Server tree, while
systemd reads the separate env file before starting the service.

**Verify:** inspect service status and logs, then open the real HTTPS App URL and
test valid and invalid login.

Publish the App and secret separately:

```bash
# Terminal — run these commands on the Shiny Server host.
sudo cp -R /srv/cerebro/apps/my_app /srv/shiny-server/
sudo chown -R shiny:shiny /srv/shiny-server/my_app/private-data/auth
sudo chmod 0500 /srv/shiny-server/my_app/private-data/auth
sudo chmod 0400 /srv/shiny-server/my_app/private-data/auth/credentials.sqlite

# Install the external secret at the exact path used by the systemd override.
sudo install -d -m 0750 /etc/cerebronexus
sudo install -m 0600 /srv/cerebro/private/viewer-auth.env /etc/cerebronexus/my_app.env
sudo chown root:root /etc/cerebronexus/my_app.env
```

Load the secret through a reviewed systemd override:

```ini
# systemd override — save this in the editor opened by `sudo systemctl edit shiny-server`.
[Service]
EnvironmentFile=/etc/cerebronexus/my_app.env
```

```bash
# Terminal — run these commands after saving the systemd override.
sudo systemctl daemon-reload
sudo systemctl restart shiny-server
```

```bash
# Terminal — run these commands to verify the Shiny Server process.
sudo systemctl status shiny-server
sudo journalctl -u shiny-server -f
```

# Deploy with Docker Compose

**Run this as:** a user allowed to run Docker.

**Run it on:** the host that contains `my-deployment/` and the external env
file.

**What happens:** Docker Compose reads `compose.yaml`, builds a runtime containing
CerebroNexus and `shinymanager`, injects the env-file assignments into the
container, and maps host port 3838 to the App.

**Verify:** open <http://localhost:3838>, test valid and invalid login, and check
`docker compose logs -f cerebro` for startup errors.

You need Docker Engine with Docker Compose. Put these files in one deployment
directory:

```text
my-deployment/
├── Dockerfile
├── compose.yaml
└── my_app/
```

The Dockerfile packages the App, not the secret:

```dockerfile
# Dockerfile — save this as `my-deployment/Dockerfile`.
FROM rocker/shiny:latest
RUN R -q -e 'install.packages(c("remotes", "shinymanager")); remotes::install_github("mihem/CerebroNexus", dependencies=TRUE)'
COPY my_app /srv/shiny-server/my_app
RUN chown -R shiny:shiny /srv/shiny-server/my_app
USER shiny
CMD ["R", "-q", "-e", "shiny::runApp('/srv/shiny-server/my_app', host='0.0.0.0', port=3838)"]
```

Save the following as `compose.yaml` beside the Dockerfile:

```yaml
# compose.yaml — save this as `my-deployment/compose.yaml` beside Dockerfile.
services:
  cerebro:
    build: .
    env_file:
      - /absolute/private/viewer-auth.env
    ports:
      - "127.0.0.1:3838:3838"
```

`env_file` is an absolute path on the host that runs Docker Compose. The file
stays outside `my-deployment/`; Compose reads it and passes the assignment into
the container environment without copying the file into the image. The port is
bound only to loopback: place a TLS-terminating reverse proxy in front of it
before exposing the Viewer to other machines.

Start the service from the deployment directory:

```bash
# Terminal — run these commands from the deployment host.
cd /path/to/my-deployment
docker compose up --build
```

Open <http://localhost:3838>. To run in the background, add `-d`:

```bash
# Terminal — run this from `my-deployment/` for background mode.
docker compose up --build -d
```

Inspect the App logs or stop the service with:

```bash
# Terminal — run these commands from `my-deployment/` to inspect or stop it.
docker compose logs -f cerebro

# Stop and remove the Compose containers and network.
docker compose down
```

Keep `viewer-auth.env` outside the build context, restrict its host permissions,
and add a defensive `.dockerignore` rule.

# Run on another machine

**Run this as:** the operator of the destination host.

**Run it on:** the new machine after the App and env file arrive through
separate encrypted transfer paths.

**What happens:** the destination R process receives the same variable name and
value used when the App database was built.

**Verify:** test the destination URL before retiring or changing the old
release.

Transfer the App and secret through separate encrypted channels, then:

```{r eval=FALSE}
# R console — run this on the destination host to start the copied App.
readRenviron("/protected/path/viewer-auth.env")
shiny::runApp("/srv/cerebronexus/my_app")
```

Validate login before retiring the old deployment.

# Limit failed sign-ins

**Run this as:** the person configuring the App startup process.

**Run it on:** every runtime host, in the R profile or startup command used by
the actual service account.

**What happens:** `shinymanager` locks an account after the configured number of
consecutive failed passwords in the current Viewer process. Restarting or
redeploying the process resets this counter from the bundled database.

**Verify:** use a disposable test account to confirm the chosen policy, then
check that the reverse proxy also performs source-aware rate limiting.

```{r eval=FALSE}
# R console — place this in the real App startup process or its `.Rprofile`.
options("shinymanager.pwd_failure_limit" = 10L)
```

Also configure source-aware rate limiting at the reverse proxy. Account locking
alone can be abused for denial of service.

# Add, remove, or rotate users

**Run this as:** the credential administrator on the trusted build host.

**Run it on:** the private source-database area, never inside a deployed App.

**What happens:** a new database and App release are created; the running App is
not updated merely because the old source database changed.

**Verify:** deploy the new App and matching secret together, test both login
outcomes, and keep the old pair only for the approved rollback window.

Create a new source database, rebuild the App, and deploy the new App with its
matching secret. Keep the old pair only for the approved rollback window.

# Troubleshooting

| Symptom | Action |
|---|---|
| No login page | Rebuild with `auth`; confirm the new App was deployed |
| Missing environment variable | Compare `passphrase_env` with the env-file name |
| Database cannot decrypt | Load the secret paired with that database |
| Deployed App stops | Inspect the service environment, not your shell |
| Database is not accessible | Give the runtime account read and directory traversal access |
| Docker rebuild loses auth | Inject the external file at runtime |

# Security checklist

- Use HTTPS.
- Keep passwords, passphrase, and env file outside the App and Git.
- Pair each App release with exactly one matching secret.
- Verify valid login, invalid login, runtime read access, and rollback.
- Add monitoring and rate limiting.

# See also

- [Create a self-contained Shiny App](create_a_self_contained_shiny_app.html)
- [Host CerebroNexus on shinyapps.io](host_cerebro_on_shinyapps.html)
- [`shinymanager::create_db()`](https://datastorm-open.github.io/shinymanager/reference/create_db.html)
