---
title: "Extra Material"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Extra Material}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

The **Extra material** tab provides a space for custom tables or plots
bundled alongside your single-cell data. Common use cases include:

- Cell type annotation results (e.g., SingleR scores)
- Custom QC summary tables
- Publication-ready figures
- Any additional analysis you want to share with collaborators

The tab appears _conditionally_ — when the `.crb` contains embedded material
or a generated app includes external tables through `createShinyApp()`. You
control what appears, so the interface stays clean when no extra content is
available.

# Quick start

```{r, eval = FALSE}
library(CerebroNexus)
launchCerebro()
```

1. Launch CerebroNexus and load a `.crb` file with extra material
2. If extra material is present, **Extra material** appears in the sidebar
3. Select a category (`tables` or `plots`)
4. Choose a specific table or plot to view

# Content categories

**Tables**: any `data.frame`. Rendered as interactive DT tables with search,
sort, and download.

**Plots**: any `ggplot2` object. Displayed at original resolution with
download support.

# Embedding content in a CRB

Use the `extra_material` parameter in `exportFromSeurat()`:

```{r, eval = FALSE}
exportFromSeurat(seurat_object,
  file = "my_data.crb",
  extra_material = list(
    tables = list(SingleR_results = annotation_df),
    plots  = list(umap_overview = umap_plot)
  )
)
```

Embedded tables and plots travel with the `.crb` and remain available wherever
that data file is opened.

# Supplying external tables to a generated app

Use `extra_tables` in `createShinyApp()` when tables should belong to one
generated deployment rather than to the CRB itself:

```{r, eval = FALSE}
createShinyApp(
  cerebro_data = c("PBMC" = "pbmc.crb"),
  result_dir = "pbmc-viewer",
  extra_tables = list(
    "QC summary" = "results/qc.csv",
    "Annotation workbook" = "results/annotations.xlsx"
  ),
  extra_tables_sheets = list(
    "Annotation workbook" = list("Cell labels" = "Sheet1")
  ),
  launch_browser = FALSE
)
```

CSV, TSV, TXT, XLS, XLSX, and XLSM files are converted during the build to
private RDS assets. Non-empty workbook sheets appear separately; mapped names
replace their source sheet names, while unmapped names are retained. The Viewer
loads a sheet only after it is selected. Original source paths are not written
to the generated configuration, and formula-like text is neutralized before
display or download.

In short, `exportFromSeurat(extra_material = ...)` embeds content in the CRB;
`createShinyApp(extra_tables = ...)` adds deployment-specific external tables
without modifying that CRB.

# See also

`vignette("export_and_visualize_custom_tables_and_plots")` for a detailed
walkthrough with examples.
