---
title: "Million-cell Viewer backend hot paths"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Million-cell Viewer backend hot paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

hot_paths <- data.frame(
  operation = c(
    "Full projection selection",
    "Filtered projection selection",
    "Cell-index resolution",
    "Single-gene expression",
    "Shuffled single-gene expression",
    "RGB expression",
    "Multi-panel expression",
    "Mean expression"
  ),
  scale = c(
    "1M cells, all groups, 100%",
    "1M cells, 17/33 clusters, 25%",
    "1M shuffled indices",
    "1 gene x 1M cells, canonical order",
    "1 gene x 100k cells, random order",
    "3 genes x 1M cells",
    "9 genes x 1M cells",
    "100 genes x 1M cells"
  ),
  pr165_ms = c(197, 329, 79, 4063, 7886, 12498, 4309, 5912),
  thin_crb_ms = c(197, 329, 79, 4063, 7886, 12498, 4309, 5912),
  latest_ms = c(31, 28, 0, 4022, 5881, 4021, 4189, 3931),
  pr165_alloc_mib = c(199.1, 166.5, 27.1, 154.1, 33.4, 461.4, 562.3, 1916.8),
  thin_crb_alloc_mib = c(199.1, 166.5, 27.1, 154.1, 33.4, 461.4, 562.3, 1916.8),
  latest_alloc_mib = c(19.1, 39.5, 0, 157.1, 22.3, 226.1, 451.0, 65.4),
  check.names = FALSE
)
hot_paths$latest_vs_pr165 <- sprintf(
  "%+.1f%%",
  (hot_paths$latest_ms / hot_paths$pr165_ms - 1) * 100
)
hot_paths$latest_vs_thin_crb <- sprintf(
  "%+.1f%%",
  (hot_paths$latest_ms / hot_paths$thin_crb_ms - 1) * 100
)
hot_paths$allocation_change_vs_both <- sprintf(
  "%+.1f%%",
  (hot_paths$latest_alloc_mib / hot_paths$thin_crb_alloc_mib - 1) * 100
)
hot_paths$latest_ms[hot_paths$operation == "Cell-index resolution"] <- NA_real_
hot_paths$latest_vs_pr165[hot_paths$operation == "Cell-index resolution"] <- ">98.7% faster"
hot_paths$latest_vs_thin_crb[hot_paths$operation == "Cell-index resolution"] <- ">98.7% faster"

crb_lifecycle <- data.frame(
  resource = c("CRB control payload (MiB)", "Write median (ms)", "Decode median (ms)", "Hydrated read median (ms)"),
  pr165 = c(26.83, 8001, 1296, 1479),
  thin_crb = c(13.213, 231, 41, 297),
  latest = c(13.215, 238, 40, 298),
  latest_vs_pr165 = c("50.7% smaller", "33.6x faster", "32.4x faster", "5.0x faster"),
  latest_vs_thin_crb = c("+0.02%", "3.0% slower", "2.4% faster", "0.3% slower"),
  check.names = FALSE
)

linked_bundle <- data.frame(
  resource = c("Bundle construction (ms)", "JSON encoding (ms)", "R bundle object (MiB)", "JSON payload (MiB)"),
  baseline = c(3473, 3625, 135.88, 62.94),
  backend = c(253, 1961, 120.62, 48.78),
  change = c("92.7% faster; 13.7x", "45.9% faster; 1.85x", "11.2% smaller", "22.5% smaller"),
  check.names = FALSE
)
```

# Purpose

This benchmark isolates the R-side paths that prepare million-cell Viewer data before JSON serialization or browser rendering. Its purpose is to answer a practical question: when a user filters a projection or switches expression colouring, how much server time and temporary R allocation can be removed without changing the CRB schema, the selected cells, or the expression values?

The comparison deliberately stops before Canvas or WebGPU rendering. It measures metadata filtering and sampling, canonical cell-index handling, BPCells expression reads, RGB and multi-panel batching, and backend-native mean expression. Browser payload and renderer work belong to the following Viewer/WebGPU layer.

# Version and data contract

| Role | Revision | Version | Contract |
|---|---|---|---|
| PR #165 | `69893a2b` | CerebroNexus 4.4.3 | Legacy RDS CRB and original Viewer backend paths |
| Thin CRB/qs2 | `27303f21` | CerebroNexus 4.5.0 | Thin CRB v1, qs2 default, hydration, and canonical BPCells cell index |
| Backend hot paths | `1ad8abc0` | CerebroNexus 4.5.1 | Same CRB contract plus lower-copy Viewer backend paths, storage-order reads, and compact Linked Views bundles |

The Thin CRB and backend stages read the same artifact under `large-examples/1m/cerebro/`. `readCerebro()` hydrates the qs2 or RDS Thin CRB, opens the unchanged BPCells sidecar, verifies the `cell_names` checksum, and restores omitted barcode and projection row names. The benchmark does not create a second data layout.

```text
large-examples/1m/
├── source/
├── seurat/
└── cerebro/
    ├── cerebro_mouse_brain_1m_thin_qs2.crb
    └── cerebro_mouse_brain_1m.bpcells/
```

# What changed

Projection filtering now returns original integer row indices directly instead of copying the complete metadata into a temporary data frame. Expression access keeps those canonical indices through `getExpressionRow()` and `getExpressionBlock()`, so a million indices are no longer converted to barcode strings and matched back to the same BPCells columns. RGB reads are batched, multi-panel extraction avoids a full transpose, and mean expression stays in the backend-native matrix representation.

The output is unchanged: sampling still uses the same seed and eligible cells, requested order is preserved, missing names still fail explicitly, and character barcode callers remain supported.

# CRB lifecycle result

The CRB lifecycle benchmark uses five alternating warm-cache rounds on the same BPCells sidecar. The 4.5.0 and 4.5.1 values come from separate runs of the same script; their small differences show run-to-run and serialized-method variation rather than a new persistence algorithm.

```{r crb-lifecycle-table}
knitr::kable(
  crb_lifecycle,
  col.names = c(
    "Resource",
    "PR #165 / 4.4.3",
    "Thin CRB / 4.5.0",
    "Backend / 4.5.1",
    "Latest vs PR #165",
    "Latest vs Thin CRB"
  ),
  row.names = FALSE,
  digits = 1
)
```

# Viewer backend result

The PR #165 and Thin CRB Viewer hot-path implementations are identical. The code between those revisions changes serialization and hydration, so their Viewer columns intentionally share the same measured baseline. The latest column was measured from backend implementation `3fb88b8a` on 2026-09-12. Each value is a warm-cache median of three repetitions; every row first verifies identical selected cells or expression values.

```{r hot-path-table}
display_hot_paths <- hot_paths
display_hot_paths$latest_ms <- ifelse(is.na(display_hot_paths$latest_ms), "<1", format(display_hot_paths$latest_ms, trim = TRUE))
knitr::kable(
  display_hot_paths[, c(
    "operation",
    "scale",
    "pr165_ms",
    "thin_crb_ms",
    "latest_ms",
    "latest_vs_pr165",
    "latest_vs_thin_crb",
    "pr165_alloc_mib",
    "thin_crb_alloc_mib",
    "latest_alloc_mib",
    "allocation_change_vs_both"
  )],
  col.names = c(
    "Operation",
    "Scale",
    "PR #165 (ms)",
    "Thin CRB (ms)",
    "Backend (ms)",
    "Latest vs PR #165",
    "Latest vs Thin CRB",
    "PR #165 allocation (MiB)",
    "Thin CRB allocation (MiB)",
    "Backend allocation (MiB)",
    "Allocation vs both"
  ),
  row.names = FALSE,
  digits = 1
)
```

Full projection selection falls from 197 to 31 ms and filtered selection from 329 to 28 ms, moving both operations toward immediate feedback. RGB expression falls from 12.50 to 4.02 seconds, saving 8.48 seconds per update. Mean expression falls from 5.91 to 3.93 seconds while cumulative R allocation drops from 1,916.8 to 65.4 MiB. A canonical-order single-gene scan remains I/O-bound and effectively unchanged; a 100,000-cell random-order stress read improves from 7.89 to 5.88 seconds after reading BPCells columns in storage order and restoring the requested output order.

One explanatory pass through the six full-scale operations, excluding index resolution and the 100,000-cell stress row, falls from 27.308 to 16.222 seconds and from 3,460.2 to 958.2 MiB of cumulative R allocations. That is 40.6% less elapsed time and 72.3% less allocation, but it is not an end-to-end Viewer timing because the operations have different UI frequencies and the total excludes transport and rendering.

# Linked Views bundle construction

The final backend audit found two repeated operations outside the expression benchmark: the full bundle calculated a second million-cell fingerprint and copied the default projection coordinates into both `projections` and `spaces`. The optimized bundle reuses the fingerprint already cached for saved views and keeps coordinates only in `projections`, from which the existing client reconstructs its lightweight space descriptor.

```{r linked-bundle-table}
knitr::kable(
  linked_bundle,
  col.names = c(
    "Resource",
    "PR #165 and Thin CRB",
    "Backend 4.5.1",
    "Change"
  ),
  row.names = FALSE,
  digits = 2
)
```

These are three-round warm medians on the same one-million-cell Thin qs2 CRB. Bundle construction falls from 3.473 seconds to 253 milliseconds, JSON encoding from 3.625 to 1.961 seconds, the R bundle from 135.88 to 120.62 MiB, and the JSON payload from 62.94 to 48.78 MiB. The benchmark normalizes only the intentionally removed duplicate fields and verifies that every remaining before/after value is equal.

The remaining 48.78 MiB is the intentional all-cell Linked Views workspace. Sampling it, loading columns on demand, or introducing typed-array transport changes Viewer behavior or its wire protocol and therefore belongs to the separate Viewer/WebGPU layer rather than this backend branch.

# Resource interpretation

Timing columns are warm-cache wall-clock medians. Allocation columns are cumulative allocations recorded by `Rprofmem`; they are not peak resident memory. The preparation host reached 8.47 GiB maximum resident memory while creating the reusable 1M artifacts, but that preparation cost is outside the interactive hot-path comparison.

| Resource | Recorded value |
|---|---:|
| Cells | 1,000,000 |
| Genes | 27,998 |
| Clusters | 33 |
| End-to-end data preparation | 1,342.12 s |
| Preparation maximum resident memory | 8.47 GiB |
| Official source H5 | 4,216,018,749 bytes |
| Reusable Seurat output | 3,778,378,648 bytes |
| CRB plus BPCells output | 3,765,141,671 bytes |
| Thin qs2 CRB control payload | 13.21 MiB |

# Reproduce the comparison

The wrapper defaults to the exact Thin CRB baseline and current checkout. It creates detached temporary worktrees, records the PR #165, Thin CRB, and backend implementation SHAs, reuses complete artifacts, and writes a tab-separated comparison table. Existing 1M data are not downloaded or regenerated when the expected files and sidecar are complete.

```{bash reproduce, eval=FALSE}
export CEREBRO_LARGE_CACHE=/path/to/cerebro-large-examples
tests/bench/run_viewer_1m_benchmark.sh 27303f21 1ad8abc0 tests/bench/scratch/backend-4.5.1
```

The output directory contains the exact comparison identity and measured rows:

```{bash inspect-output, eval=FALSE}
column -t -s $'\t' tests/bench/scratch/backend-4.5.1/run_manifest.tsv
column -t -s $'\t' tests/bench/scratch/backend-4.5.1/hot_paths.tsv
column -t -s $'\t' tests/bench/scratch/backend-4.5.1/bundle.tsv
```

To reuse a prepared CRB directly, run the lower-level benchmark with explicit source roots. The CRB may use qs2 or RDS because the script loads it through `readCerebro()`.

```{bash lower-level, eval=FALSE}
Rscript tests/bench/viewer_1m_hot_paths.R /path/to/thin-crb-source /path/to/backend-source /path/to/cerebro_mouse_brain_1m_thin_qs2.crb 3
```

The current benchmark records a separate barcode-to-index row before the expression workloads. This exposes the work removed by retaining canonical indices instead of hiding it inside a BPCells read.

# Evidence policy

Published values must come from `hot_paths.tsv` and `bundle.tsv` together with `run_manifest.tsv`; revision labels copied by hand are insufficient. A row is accepted only when its before/after correctness check passes. Cold-cache results, browser transfer, JSON parsing, Canvas/WebGPU rendering, GPU memory, and end-to-end interaction latency are intentionally outside this backend-only benchmark and must not be inferred from it.
