---
title: "Reproduce the one-million-cell CRB artifacts"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Reproduce the one-million-cell CRB artifacts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

run_full <- identical(
  tolower(Sys.getenv("CEREBRO_RUN_1M_VIGNETTE", unset = "false")),
  "true"
)
cache_root <- Sys.getenv(
  "CEREBRO_LARGE_CACHE",
  unset = file.path(
    tools::R_user_dir("CerebroNexus", "cache"),
    "large-examples"
  )
)
data_root <- file.path(cache_root, "1m")
repo_root <- normalizePath(
  Sys.getenv("CEREBRO_REPO_ROOT", unset = ".."),
  mustWork = TRUE
)
paths <- list(
  source_h5 = file.path(
    data_root,
    "source",
    "1M_neurons_filtered_gene_bc_matrices_h5.h5"
  ),
  seurat_rds = file.path(data_root, "seurat", "mouse_brain_1m.rds"),
  seurat_bpcells = file.path(
    data_root,
    "seurat",
    "mouse_brain_1m.bpcells"
  ),
  source_crb = file.path(
    data_root,
    "cerebro",
    "cerebro_mouse_brain_1m.crb"
  ),
  cerebro_bpcells = file.path(
    data_root,
    "cerebro",
    "cerebro_mouse_brain_1m.bpcells"
  ),
  thin_rds = file.path(
    data_root,
    "cerebro",
    "cerebro_mouse_brain_1m_thin_rds.crb"
  ),
  thin_qs2 = file.path(
    data_root,
    "cerebro",
    "cerebro_mouse_brain_1m_thin_qs2.crb"
  )
)

path_size <- function(path) {
  if (file.exists(path) && !dir.exists(path)) {
    return(unname(file.info(path)$size))
  }
  if (!dir.exists(path)) {
    return(NA_real_)
  }
  files <- list.files(path, recursive = TRUE, full.names = TRUE)
  sum(file.info(files)$size, na.rm = TRUE)
}

detect_crb_codec <- function(path) {
  if (!file.exists(path) || dir.exists(path)) {
    return("RDS or qs2 (auto-detected)")
  }
  connection <- file(path, open = "rb")
  on.exit(close(connection), add = TRUE)
  magic <- readBin(connection, "raw", n = 4L)
  if (identical(magic, as.raw(c(0x0b, 0x0e, 0x0a, 0xc1)))) {
    "qs2"
  } else {
    "RDS"
  }
}

artifact_table <- function() {
  bytes <- vapply(paths, path_size, numeric(1))
  data.frame(
    stage = c(
      "Public source",
      "Processed Seurat object",
      "Seurat expression sidecar",
      "Input CRB",
      "Cerebro expression sidecar",
      "Thin compatibility CRB",
      "Thin default CRB"
    ),
    file_name = basename(unlist(paths, use.names = FALSE)),
    logical_format = c(
      "10x feature-barcode matrix",
      "Seurat",
      "Expression matrix",
      "Cerebro CRB",
      "Expression matrix",
      "Cerebro CRB",
      "Cerebro CRB"
    ),
    physical_format = c(
      "HDF5",
      "RDS",
      "BPCells directory (uint32)",
      detect_crb_codec(paths$source_crb),
      "BPCells directory (uint32)",
      "RDS",
      "qs2"
    ),
    exists = vapply(
      paths,
      function(path) file.exists(path) || dir.exists(path),
      logical(1)
    ),
    size_gib = ifelse(is.na(bytes), NA, round(bytes / 1024^3, 3)),
    relative_path = substring(unlist(paths, use.names = FALSE), nchar(data_root) + 2L),
    check.names = FALSE
  )
}
```

# Purpose

This vignette records exactly which public data are used for the one-million-cell CRB benchmark, how the data are transformed, where every artifact is stored, and how to reproduce both physical CRB codecs. It measures storage and I/O engineering behavior; the deterministic one-million-cell subset is not a new biological analysis cohort.

# Source data

The source is the 10x Genomics **1M Neurons** data set: E18 mouse brain nuclei processed by Cell Ranger 1.3.0. The public HDF5 matrix contains 1,306,127 cell barcodes and 27,998 features.

```text
https://cf.10xgenomics.com/samples/cell-exp/1.3.0/1M_neurons/
  1M_neurons_filtered_gene_bc_matrices_h5.h5
```

The preparation script downloads this file without modifying it. The generated data do **not** live under the repository root. The repository contains the preparation and benchmark code, while the large source and derived artifacts live in an external user-cache directory.

| Location | Resolved path | Relationship |
|---|---|---|
| Repository root | `r repo_root` | Git-tracked source, tests, and vignette only |
| Large-example cache root | `r normalizePath(cache_root, mustWork = FALSE)` | External to the repository; replaces `<cache-root>` below |
| One-million-cell data root | `r normalizePath(data_root, mustWork = FALSE)` | `<cache-root>/1m/` |

By default, `<cache-root>` is `file.path(tools::R_user_dir("CerebroNexus", "cache"), "large-examples")`. On the recorded macOS host this resolves to `/Users/nuioi/Library/Caches/org.R-project.R/R/CerebroNexus/large-examples/`. Set `CEREBRO_LARGE_CACHE` to replace that complete cache-root path; for example, `CEREBRO_LARGE_CACHE=/Volumes/FastSSD/cerebro-large-examples` stores the dataset at `/Volumes/FastSSD/cerebro-large-examples/1m/`. A fresh run needs approximately 16 GiB of free space.

# Processing steps

The transformation is deterministic:

1. Open the public 10x HDF5 matrix with BPCells.
2. Select exactly 1,000,000 barcodes at evenly spaced source indexes.
3. Replace Ensembl row identifiers with unique `gene_names` symbols.
4. Convert non-negative integer counts to BPCells `uint32_t` storage.
5. Run `NormalizeData()` and `FindVariableFeatures()`.
6. Draw a 50,000-cell leverage-score sketch with seed 123.
7. Run 30-component PCA, neighbours, clustering at resolution 1, and UMAP with seed 123 on the sketch.
8. Project the PCA, cluster labels, and UMAP back to all 1,000,000 cells.
9. Save a Seurat RDS plus its BPCells expression sidecar.
10. Export counts, `sample`, `seurat_clusters`, and the full UMAP to a BPCells-backed CRB.
11. Re-encode only the CRB control payload as thin RDS and thin qs2. Both files reuse the same unchanged Cerebro BPCells sidecar.

The thin CRB uses the sidecar's ordered `col_names` as its canonical cell index. It omits the live expression handle, `meta_data$cell_barcode`, and projection barcode row names, records the `col_names` checksum and portable-view cell fingerprint, and restores the omitted fields in `readCerebro()`. Computing the order-independent fingerprint once during export avoids sorting and hashing one million barcodes whenever the Viewer starts; older CRBs remain readable and calculate it at runtime.

# Files and formats

The following table is generated by the code in this vignette. `exists` and `size_gib` reflect the current cache; the file names, formats, and relative locations are fixed by the preparation workflow.

```{r artifact-manifest}
knitr::kable(artifact_table(), row.names = FALSE)
```

The resulting directory layout is relative to the external `<cache-root>`, not to `<repository-root>`:

```text
<cache-root>/
└── 1m/
    ├── source/
    │   └── 1M_neurons_filtered_gene_bc_matrices_h5.h5
    ├── seurat/
    │   ├── mouse_brain_1m.rds
    │   └── mouse_brain_1m.bpcells/
    └── cerebro/
        ├── cerebro_mouse_brain_1m.crb
        ├── cerebro_mouse_brain_1m.bpcells/
        ├── cerebro_mouse_brain_1m_thin_rds.crb
        └── cerebro_mouse_brain_1m_thin_qs2.crb
```

`.crb` identifies the logical Cerebro data format, not the physical codec. Use `readCerebro()` for every new or old CRB. Direct `readRDS()` remains valid only for files explicitly written with `codec = "rds"` and legacy RDS CRBs.

# Comparison against PR #165

The benchmark baseline is the legacy RDS CRB from PR #165 (`69893a2b`, CerebroNexus 4.4.3). The optimization branch is based directly on that merge commit and develops the CRB changes as CerebroNexus 4.5.0.

## Recorded environment

| Item | Recorded value |
|---|---|
| Machine | MacBook Pro (`MacBookPro18,1`) |
| CPU | Apple M1 Pro, 10 cores (8 performance + 2 efficiency) |
| Memory | 32 GiB |
| Storage | Apple SSD `AP1024R`, 1 TB, APFS, Apple Fabric |
| Operating system | macOS 27.0, build `26A5421a` |
| Architecture | `arm64` |
| R | 4.6.1 (`aarch64-apple-darwin23`) |
| BPCells | 0.3.1 |
| qs2 | 0.2.2 |
| Seurat / SeuratObject | 5.5.1 / 5.4.0 |
| Benchmark baseline | PR #165 (`69893a2b`), CerebroNexus 4.4.3 |
| Optimization candidate | `perf/pr0-million-cell-crb-qs2`, based on PR #165 (`69893a2b`), CerebroNexus 4.5.0 development version |

Measurements are warm-cache medians from five alternating rounds on the same unchanged BPCells sidecar. The benchmark loads the source checkout with `pkgload::load_all()`; the separately installed CerebroNexus 5.0.0 package is not used.

`Write` includes thin-payload compaction and serialization. `Decode` measures only physical deserialization. `Hydrated read` measures `readCerebro()`, including reopening the BPCells sidecar, validating its canonical cell index checksum, and restoring omitted metadata and projection row names.

```{r benchmark-data, include=FALSE}
benchmark_results <- data.frame(
  candidate = c(
    "PR #165 legacy CRB (69893a2b, v4.4.3)",
    "Thin CRB without qs2 (v4.5.0-dev)",
    "Thin CRB with qs2 (v4.5.0-dev)"
  ),
  codec = c("RDS", "RDS", "qs2"),
  size_mib = c(26.8290309906006, 13.9180183410645, 13.2125005722046),
  write_median_ms = c(8001, 2947, 231),
  decode_median_ms = c(1296, 93, 41),
  hydrated_median_ms = c(1479, 356, 297)
)

baseline <- benchmark_results[1, ]
benchmark_results$size_reduction_pct <- 100 * (1 - benchmark_results$size_mib / baseline$size_mib)
benchmark_results$write_speedup <- baseline$write_median_ms / benchmark_results$write_median_ms
benchmark_results$decode_speedup <- baseline$decode_median_ms / benchmark_results$decode_median_ms
benchmark_results$hydrated_speedup <- baseline$hydrated_median_ms / benchmark_results$hydrated_median_ms
```

| Candidate | Codec | CRB size | Write median | Decode median | Hydrated read median | Size reduction vs PR #165 | Write speedup | Decode speedup | Hydrated read speedup |
|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| PR #165 legacy CRB (`69893a2b`, v4.4.3) | RDS | `r sprintf("%.2f MiB", benchmark_results$size_mib[1])` | `r sprintf("%.0f ms", benchmark_results$write_median_ms[1])` | `r sprintf("%.0f ms", benchmark_results$decode_median_ms[1])` | `r sprintf("%.0f ms", benchmark_results$hydrated_median_ms[1])` | — | — | — | — |
| Thin CRB without qs2 (v4.5.0-dev) | RDS | `r sprintf("%.2f MiB", benchmark_results$size_mib[2])` | `r sprintf("%.0f ms", benchmark_results$write_median_ms[2])` | `r sprintf("%.0f ms", benchmark_results$decode_median_ms[2])` | `r sprintf("%.0f ms", benchmark_results$hydrated_median_ms[2])` | `r sprintf("%.2f%%", benchmark_results$size_reduction_pct[2])` | `r sprintf("%.2fx", benchmark_results$write_speedup[2])` | `r sprintf("%.2fx", benchmark_results$decode_speedup[2])` | `r sprintf("%.2fx", benchmark_results$hydrated_speedup[2])` |
| Thin CRB with qs2 (v4.5.0-dev) | qs2 | `r sprintf("%.2f MiB", benchmark_results$size_mib[3])` | `r sprintf("%.0f ms", benchmark_results$write_median_ms[3])` | `r sprintf("%.0f ms", benchmark_results$decode_median_ms[3])` | `r sprintf("%.0f ms", benchmark_results$hydrated_median_ms[3])` | `r sprintf("%.2f%%", benchmark_results$size_reduction_pct[3])` | `r sprintf("%.2fx", benchmark_results$write_speedup[3])` | `r sprintf("%.2fx", benchmark_results$decode_speedup[3])` | `r sprintf("%.2fx", benchmark_results$hydrated_speedup[3])` |

Without qs2, thin CRB reduces the control file by 48.12%, accelerates writing by 2.71x, physical decoding by 13.94x, and hydrated `readCerebro()` by 4.15x relative to PR #165. With qs2, the reductions are 50.75%, 34.64x, 31.61x, and 4.98x respectively. The incremental file-size benefit of qs2 over thin RDS is only 5.07%; most of the size reduction therefore comes from removing redundant CRB fields rather than changing codec.

The checksum failure covered by the automated test is deliberately injected to prove that a CRB cannot silently attach the wrong BPCells sidecar. The genuine one-million-cell RDS and qs2 files used here both pass checksum validation.

The following command reruns all five alternating benchmark rounds and regenerates the source CSV used to record these measurements:

```{bash, eval=FALSE}
BASELINE=$(Rscript -e 'cat(file.path(tools::R_user_dir("CerebroNexus", "cache"), "large-examples", "1m", "cerebro", "cerebro_mouse_brain_1m.crb"))')
Rscript tests/bench/benchmark_million_cell_crb.R "$BASELINE" 5 tests/bench/results/million_cell_crb_pr165.csv
```

# Run the complete preparation

The expensive chunks are disabled during ordinary package builds. Run the complete workflow from the repository root with:

```{bash, eval=FALSE}
CEREBRO_RUN_1M_VIGNETTE=true \
Rscript -e 'rmarkdown::render("vignettes/million_cell_crb_artifacts.Rmd")'
```

The following code performs the actual download, analysis, export, and codec conversion when `CEREBRO_RUN_1M_VIGNETTE=true`.

```{r prepare-one-million, eval=run_full, message=TRUE}
source(file.path(repo_root, "tests", "bench", "prepare_viewer_1m_data.R"))
pkgload::load_all(repo_root, quiet = TRUE)

source_crb <- prepareViewer1mBenchmarkData(cache_dir = cache_root)
object <- readCerebro(source_crb)
stopifnot(length(object$getCellNames()) == 1000000L)

saveCerebro(object, paths$thin_rds, codec = "rds")
saveCerebro(object, paths$thin_qs2, codec = "qs2")
```

# Verify the outputs and regenerate the table

This final chunk reads both physical codecs through the same public API, checks their exact cell and UMAP ordering, and regenerates the artifact table from the files on disk.

```{r verify-and-tabulate, eval=run_full}
expected_cells <- object$getCellNames()
expected_umap_cells <- rownames(object$getProjection("umap"))

for (path in c(paths$thin_rds, paths$thin_qs2)) {
  restored <- readCerebro(path)
  stopifnot(
    identical(restored$getCellNames(), expected_cells),
    identical(rownames(restored$getProjection("umap")), expected_umap_cells),
    identical(restored$meta_data$cell_barcode, expected_cells)
  )
  rm(restored)
  gc()
}

knitr::kable(artifact_table(), row.names = FALSE)
```

The BPCells directory is intentionally not copied for each codec. The RDS and qs2 `.crb` files contain alternative serializations of the same Cerebro control data and both point to `cerebro_mouse_brain_1m.bpcells/`.
