---
title: "Benchmark the million-cell Viewer from PR0 through PR4"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Benchmark the million-cell Viewer from PR0 through PR4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Purpose

This guide reproduces the million-cell workload used to profile CerebroNexus from PR0 through PR4. It covers five separate operations:

1. download the official 10x Genomics E18 mouse-brain matrix;
2. build a reproducible one-million-cell Seurat object and BPCells-backed CRB;
3. compare the PR0 compact CRB, PR1 backend hot paths, PR2 WebGPU Viewer, and PR3 startup changes;
4. measure the PR4 gene-major expression layout and slow-page follow-up;
5. launch the same data interactively through the bundled demo app.

The expensive chunks are deliberately not evaluated while building the vignette. A package build must never download a 3.93 GiB source matrix or start a browser benchmark.

# PR0–PR4 optimization line

The five PRs change different layers and therefore use separate benchmark harnesses. CRB lifecycle, R backend operations, installed-package startup, browser navigation, and renderer-only measurements must not be combined into one synthetic end-to-end number.

| PR | Measurement point | Optimization layer | Main change |
|---|---|---|---|
| PR0 | `27303f21` | CRB persistence | Thin control payload, qs2 serialization, BPCells hydration |
| PR1 | `1ad8abc0` | Viewer backend | Integer indexes, batched/storage-order reads, backend-native aggregation, compact bundles |
| PR2 | `010be5fe` / `6b7dce4a` | Browser rendering | Binary transport, WebGPU, deferred hover and first-frame work |
| PR3 | `639acb89` | Installed startup | Thin CRB schema v2, stored cell count, lazy hydration, deferred inactive servers |
| PR4 | Working tree after `f5fc8bc9` | Expression layout and slow pages | Gene-major BPCells export, cached page state, HLA Canvas, Trajectory first-frame scheduling |

The headline measurements below use the official 10x E18 mouse brain at 1,000,000 cells and 27,998 genes. PR4 page measurements additionally use a deterministic fixture containing 1,000,000 trajectory rows, 100,000 synthetic receptor rows, and 32 synthetic HLA-typed samples.

| PR | Representative metric | Before | After | Change |
|---|---|---:|---:|---:|
| PR0 | CRB control payload | 26.829 MiB | 13.213 MiB | 50.7% smaller |
| PR0 | CRB write median | 8,001 ms | 231 ms | 34.6x faster |
| PR0 | CRB decode median | 1,296 ms | 41 ms | 31.6x faster |
| PR1 | Full projection selection | 197 ms | 31 ms | 84.3% faster |
| PR1 | RGB, 3 genes × 1M cells | 12,498 ms | 4,021 ms | 67.8% faster |
| PR1 | Mean, 100 genes × 1M cells | 5,912 ms | 3,931 ms | 33.5% faster |
| PR2 | First Overview, 1M cells | 48.768 s Canvas | 1.377 s WebGPU | 97.2% faster |
| PR2 | Zoom median, 1M cells | 280.9 ms Canvas | 8.0 ms WebGPU | 35.1x faster |
| PR3 | Process start to Data Info | 3.692 s schema v1 | 2.557 s schema v2 | 30.7% faster |
| PR3 | Process start to Data Info vs Legacy RDS | 9.045 s | 2.557 s | 71.7% faster cumulatively |
| PR4 | RGB, 3 genes × 1M cells | 3,984 ms column-major | 202 ms gene-major | 94.9% faster |
| PR4 | Mean, 100 genes × 1M cells | 3,748 ms column-major | 55 ms gene-major | 98.5% faster |
| PR4 | Trajectory first visit | 3,660 ms PR3 | 2,434 ms PR4 | 33.5% faster |
| PR4 | HLA & TCR Motifs first frame | 5,913 ms PR3 | 2,696 ms PR4 | 54.4% faster, directional |

The PR4 HLA result waits for every Canvas node and edge to be painted. The older PR3 measurement waited for the broader page-level Shiny idle state, so the HLA percentage shows the practical direction and target result rather than a publication-grade same-harness estimate. The detailed sections below preserve each benchmark's own scope and provenance.

# Requirements and disk use

Run the commands from the root of a CerebroNexus checkout. Data preparation
requires `BPCells`, `Seurat`, `SeuratObject`, and `rhdf5`. The browser benchmark
also requires `shinytest2` and a Chromium build with WebGPU support.

The source used here contains 27,998 genes and 1,306,127 cells. The preparation
script selects exactly 1,000,000 cells at evenly spaced indexes, replaces
Ensembl identifiers with unique gene symbols, and computes a sketch-based UMAP
with seed 123.

Allow at least 16 GiB of free disk space for the source file, the Seurat and CRB
BPCells directories, and transactional staging during export. The default cache
is below `tools::R_user_dir("CerebroNexus", "cache")`; pass `cache_dir` to put it
on a larger volume.

# Download and prepare the data

The preparation script downloads the official 10x HDF5 matrix only when it is
absent, then reuses every complete intermediate on later runs.

```{r prepare-data}
devtools::load_all(quiet = TRUE)
source("tests/bench/prepare_viewer_1m_data.R")

crb <- prepareViewer1mBenchmarkData()
crb
```

To choose another cache root:

```{r prepare-data-custom-cache}
crb <- prepareViewer1mBenchmarkData(
  cache_dir = "/Volumes/large-data/CerebroNexus-benchmark"
)
```

The original download is:

```text
https://cf.10xgenomics.com/samples/cell-exp/1.3.0/1M_neurons/1M_neurons_filtered_gene_bc_matrices_h5.h5
```

The resulting cache contains:

```text
1m/
├── source/1M_neurons_filtered_gene_bc_matrices_h5.h5
├── seurat/mouse_brain_1m.rds
├── seurat/mouse_brain_1m.bpcells/
├── cerebro/cerebro_mouse_brain_1m.crb
└── cerebro/cerebro_mouse_brain_1m.bpcells/
```

The `.crb` and adjacent `.bpcells/` directory are one portable dataset. Keep
them together when moving or deploying it.

# Run the regression tests

The focused regression set covers expression export and integrity, the compact
BPCells CRB contract, the million-cell payload, and renderer integration.

```{bash regression-tests, eval=FALSE}
Rscript -e 'devtools::test(
  filter = paste0(
    "million-cell-renderer|million-cell-viewer|",
    "seurat-v5-split-layers|export-data-integrity|",
    "exportFromSeurat|r-functions"
  ),
  stop_on_failure = TRUE
)'
```

The real BPCells bundle relocation check can be run separately:

```{bash bundle-test, eval=FALSE}
Rscript -e 'devtools::load_all(quiet = TRUE); testthat::test_file(
  "tests/testthat/test-createShinyApp-sibling.R",
  desc = "a bundled real BPCells backend attaches with exact data",
  stop_on_failure = TRUE
)'
```

On 12 September 2026 these commands produced 308 passes, zero failures, zero
warnings, and one conditional HDF5 skip, followed by 7/7 passes for the real
BPCells bundle check.

# Compare the Viewer before and after

## Create the baseline checkout

Commit `c455d0cd` is the exact 4.5.1 baseline before the WebGPU work. It has no one-million-cell menu entry and still uses the Canvas 2D renderer, but the benchmark launches its `launchCerebro()` directly against the same CRB. The fixed revision keeps the comparison reproducible after development branches are merged or removed.

| Candidate | Revision | Version | Role in this comparison |
|---|---|---|---|
| Canvas baseline | `c455d0cd` | 4.5.1 | Optimized CRB/backend, Canvas 2D Viewer, no bundled 1M demo |
| WebGPU release | This source tree | 4.6.0 | Binary transport, WebGPU rendering, deferred hover payload, bundled opt-in 1M demo |

Create a detached worktree so both implementations can be measured against the same CRB and browser installation:

```{bash baseline-worktree, eval=FALSE}
git worktree add --detach ../CerebroNexus-4.5.1-baseline c455d0cd
```

Use an old-format BPCells CRB for this Viewer comparison. The current Viewer
loads that format too, so both candidates receive exactly the same serialized
input. This keeps the CRB-format reduction separate from browser rendering.

## Run the full Viewer benchmark

The 4.5.1-to-4.6.0 comparison uses the harness's Overview-only mode because the 4.5.1 Canvas implementation cannot provide a usable one-million-cell Linked Views, gene-expression, or RGB workflow. The harness still checks that Data Info reports 1,000,000 cells, verifies that every expected point was rendered, performs ten alternating zoom operations, alternates candidate order on each round, and rejects GPU errors or context loss. The later WebGPU-to-WebGPU comparison exercises every page path.

Replace `CRB` below with the path returned by the preparation step.

```{bash viewer-benchmark, eval=FALSE}
CRB=/path/to/cerebro_mouse_brain_1m.crb

CEREBRO_VIEWER_BENCH_OVERVIEW_ONLY=1 \
  Rscript tests/bench/benchmark_million_cell_viewer.R \
  ../CerebroNexus-4.5.1-baseline . "$CRB" 3 100

CEREBRO_VIEWER_BENCH_OVERVIEW_ONLY=1 \
  Rscript tests/bench/benchmark_million_cell_viewer.R \
  ../CerebroNexus-4.5.1-baseline . "$CRB" 3 10
```

The 100% run draws all one million cells. The 10% run draws 100,000 cells and
matches the bundled demo's initial setting. Timings below are medians of three
fresh app processes per candidate on the same host.

| Initial display | Metric | Before (`canvas2d`) | After (`webgpu`) | Change |
|---|---|---:|---:|---:|
| 100% | Data ready | 15.088 s | 16.313 s | +1.225 s in development harness |
| 100% | First Overview | 48.768 s | 1.377 s | 97.2% faster |
| 100% | Data + Overview | 63.856 s | 17.690 s | 72.3% faster |
| 100% | Zoom median | 280.9 ms | 8.0 ms | 35.1x faster |
| 100% | Zoom p95 | 310.3 ms | 15.4 ms | 20.1x faster |
| 10% | Data ready | 14.836 s | 15.613 s | +0.777 s in development harness |
| 10% | First Overview | 35.838 s | 0.323 s | 99.1% faster |
| 10% | Data + Overview | 50.674 s | 15.936 s | 68.6% faster |
| 10% | Zoom median | 91.5 ms | 3.7 ms | 24.7x faster |
| 10% | Zoom p95 | 108.7 ms | 5.9 ms | 18.4x faster |

The first row group shows why storage and rendering are reported separately: loading and attaching the same CRB barely changed, while the first plot and subsequent camera operations changed substantially. The missing 4.5.1 Linked Views and expression rows are intentional: at one million points those Canvas paths are the unusable baseline this release replaces, not a workflow worth presenting as a demo.

## Profile cold startup

The `Data ready` rows above are end-to-end development-harness measurements, not installed-package startup measurements. The timer starts before `shinytest2::AppDriver$new()`, and the generated child application begins with `devtools::load_all(root)`. A sampled 16.93-second source-tree run attributed about 6.15 seconds to `load_all()`, 2.96 seconds to reading the legacy RDS CRB, 1.63 seconds to sorting and hashing one million cell names, 0.78 seconds to the intentional Linked Views prewarm, 0.10 seconds to constructing the app, and less than 0.10 seconds to the Data Info output itself. Sampled stacks can overlap and do not sum to wall time, but they identify the expensive work.

This also explains why the earlier CRB and backend passes did not expose one 15-second block. `benchmark_million_cell_crb.R` loads the package and warms every file before timing write, decode, and hydrated read operations. `viewer_1m_hot_paths.R` loads the package, reads the CRB, and materializes metadata before its `elapsed_ms()` loops. Those benchmarks correctly isolated their target operations, but neither measured cold Viewer startup.

Use the installed-package harness below to measure startup without charging source loading to the product. It records package attachment, app construction, server listening, browser-to-data readiness, and total process-to-data readiness separately. Candidate order reverses on alternate rounds.

```{bash startup-benchmark, eval=FALSE}
STARTUP_LIB=$(mktemp -d /tmp/cerebronexus-startup.XXXXXX)
R CMD INSTALL --library="$STARTUP_LIB" .

R_LIBS="$STARTUP_LIB" \
  CEREBRO_STARTUP_REPEATS=3 \
  CEREBRO_STARTUP_OUTPUT=tests/bench/results/million_cell_startup_4_6_0.csv \
  Rscript tests/bench/benchmark_million_cell_startup.R \
  legacy_rds=/path/to/cerebro_mouse_brain_1m.crb \
  thin_qs2_no_fingerprint=/path/to/cerebro_mouse_brain_1m_thin_qs2.crb \
  thin_qs2_fingerprint=/path/to/cerebro_mouse_brain_1m_fingerprint_qs2.crb
```

The recorded values are medians of three fresh Viewer processes per artifact. All three candidates use the same installed 4.6.0 code from before the runtime scheduling optimization and an unchanged BPCells sidecar, so this comparison isolates the control-file representation and startup fingerprint work.

| CRB artifact | Package attach | App construct | Server listening | Browser to data | Process to data |
|---|---:|---:|---:|---:|---:|
| Legacy RDS | 1.010 s | 0.192 s | 1.384 s | 7.558 s | 9.045 s |
| Thin qs2 without stored fingerprint | 0.999 s | 0.192 s | 1.381 s | 6.490 s | 7.977 s |
| Thin qs2 with stored fingerprint | 1.015 s | 0.193 s | 1.382 s | 4.559 s | 6.024 s |

Thin qs2 removes 11.8% of legacy process-to-data time. Recording the fingerprint during export brings the total reduction to 33.4% against legacy RDS and 24.5% against the previous thin qs2 artifact.

The runtime optimization was then measured separately against its exact code starting point with the fingerprinted thin qs2 artifact, the same BPCells sidecar, and three fresh installed-package processes per candidate. The startup harness now records browser document loading separately from the interval between the load event and Data Info becoming ready. Raw measurements are in `tests/bench/results/million_cell_startup_optimization_4_6_0.csv`.

| Startup phase | Starting point | Optimized | Change |
|---|---:|---:|---:|
| Package attach | 1.023 s | 1.025 s | 0.2% slower |
| App construction | 0.192 s | 0.203 s | 5.7% slower |
| Process start to server listening | 1.387 s | 1.432 s | 3.3% slower |
| Browser document load | 0.292 s | 0.260 s | 10.9% faster |
| Load event to Data ready | 4.291 s | 2.287 s | 46.7% faster |
| Browser navigation to Data ready | 4.589 s | 2.543 s | 44.6% faster |
| Process start to Data ready | 6.081 s | 4.097 s | 32.6% faster |

The optimized thin BPCells path validates and hydrates the million cell names without loading BPCells, then attaches the expression matrix on its first real use. Large-dataset Linked Views prewarming starts after the initial response and a three-second grace period; its gene-selector gate checks that grace period before inspecting the expression matrix. Overview and Data Info therefore no longer wait for a hidden expression control or hidden Linked Views bundle. The cost is deferred rather than erased: opening an expression-dependent page immediately performs the one-time BPCells attach, while an otherwise idle session performs it during Linked Views prewarming.

PR3 adds Thin CRB schema v2 with a stored cell count, allowing Data Info to report the true one-million-cell size without first parsing the sidecar cell index or reconstructing metadata and projection row names. The Viewer registers optional page servers 100 ms after the first response and resumes collapsed group-filter outputs after 500 ms; package-level `readCerebro()` remains eager, old schema v1 files remain compatible, and browser regressions cover the deferred pages. The strict gate is the median of five fresh installed-package processes and requires `process_to_data_ms < 3000` without changing the Data Info readiness selector. Raw measurements are in `tests/bench/results/million_cell_startup_sub3_4_6_0.csv`.

```{bash pr3-startup-gate, eval=FALSE}
R_LIBS="$STARTUP_LIB" \
  CEREBRO_STARTUP_REPEATS=5 \
  CEREBRO_STARTUP_GATE_LABEL=v2_lazy \
  CEREBRO_STARTUP_MAX_PROCESS_TO_DATA_MS=3000 \
  CEREBRO_STARTUP_OUTPUT=tests/bench/results/million_cell_startup_sub3_4_6_0.csv \
  Rscript tests/bench/benchmark_million_cell_startup.R \
  v1_eager=/path/to/cerebro_mouse_brain_1m_thin_qs2.crb \
  v2_lazy=/path/to/cerebro_mouse_brain_1m_startup_v2.crb
```

| Startup phase | Schema v1 eager | Schema v2 lazy | Change |
|---|---:|---:|---:|
| Package attach | 1.051 s | 1.048 s | 0.3% faster |
| App construction | 0.199 s | 0.201 s | 0.7% slower |
| Process start to server listening | 1.434 s | 1.433 s | 0.1% faster |
| Browser document load | 0.315 s | 0.296 s | 6.2% faster |
| Load event to Data Info | 1.770 s | 0.733 s | 58.6% faster |
| Browser navigation to Data Info | 2.105 s | 1.029 s | 51.1% faster |
| Process start to Data Info | 3.692 s | 2.557 s | 30.7% faster |

The 2.557-second median passes the `<3 s` gate with 443 ms of headroom. It is 37.6% faster than the preceding 4.097-second installed-runtime optimization point and 71.7% faster than the original 9.045-second Legacy RDS path.

New BPCells CRBs written by `saveCerebro()` or an exporter contain the fingerprint automatically. To make an optimized artifact from an older thin CRB while preserving the original benchmark input, write the copy beside the same BPCells sidecar:

```{r upgrade-crb}
old_crb <- "/path/to/cerebro_mouse_brain_1m_thin_qs2.crb"
current_crb <- file.path(dirname(old_crb), "cerebro_mouse_brain_1m_startup_v2.crb")
saveCerebro(readCerebro(old_crb), current_crb)
```

## Verify the three-second navigation target

The later hot-path pass starts from commit `5a915962`, where WebGPU was already
enabled. This comparison therefore measures transport, first-frame scheduling,
group-label calculation, and large-dataset bundle preparation without counting
the earlier Canvas-to-WebGPU replacement as a second time.

```{bash three-second-baseline, eval=FALSE}
git worktree add --detach ../CerebroNexus-before-wire 5a915962

Rscript tests/bench/benchmark_million_cell_viewer.R \
  ../CerebroNexus-before-wire . "$CRB" 3 100

Rscript tests/bench/benchmark_million_cell_viewer.R \
  ../CerebroNexus-before-wire . "$CRB" 3 10
```

Both candidates used WebGPU. Results are medians of three alternating fresh processes on 12 September 2026. The selector is stabilized before timing so the expression rows include the actual million-cell BPCells read rather than a stale render:

| Display | Metric | Before | After | Change |
|---|---|---:|---:|---:|
| 100% (1,000,000 points) | Data ready | 15.455 s | 16.330 s | +0.875 s in development harness |
| 100% (1,000,000 points) | First Overview | 11.481 s | 1.440 s | 87.5% faster |
| 100% (1,000,000 points) | Linked Views ready | 4.462 s | 2.507 s | 43.8% faster |
| 100% (1,000,000 points) | Gene expression ready | 88.790 s | 81.846 s | 7.8% faster |
| 100% (1,000,000 points) | RGB ready | 99.863 s | 82.157 s | 17.7% faster |
| 100% (1,000,000 points) | Zoom median | 8.1 ms | 8.4 ms | effectively tied |
| 100% (1,000,000 points) | Zoom p95 | 15.1 ms | 22.0 ms | 6.9 ms slower |
| 10% (100,000 points) | Data ready | 13.693 s | 14.081 s | +0.388 s in development harness |
| 10% (100,000 points) | First Overview | 2.582 s | 1.526 s | 40.9% faster |
| 10% (100,000 points) | Linked Views ready | 4.942 s | 1.187 s | 76.0% faster |
| 10% (100,000 points) | Gene expression ready | 1.018 s | 0.869 s | 14.6% faster |
| 10% (100,000 points) | RGB ready | 1.795 s | 1.470 s | 18.1% faster |
| 10% (100,000 points) | Zoom median | 4.6 ms | 3.8 ms | 17.4% faster |
| 10% (100,000 points) | Zoom p95 | 6.6 ms | 5.3 ms | 19.7% faster |

Overview and Linked Views now meet the three-second navigation target. Integer vectors use the narrowest lossless wire type, reducing this dataset's initial Linked Views bundle from 28,001,836 to 15,001,832 bytes (46.4% smaller). Hover columns remain in their grouped representation and are expanded only for the point actually under the pointer. Non-render-critical hover data is sent after the first-frame work rather than competing with it.

At the PR2 stage, the full-million gene and RGB rows remained dominated by reading expression values from the 3.5 GiB BPCells sidecar. WebGPU kept their eventual million-point render responsive and avoided browser lock-up, but it could not accelerate the storage scan. PR4 addresses that separate storage problem with a gene-major BPCells layout.

The 14-second `Data ready` values in this table use the same source-tree harness described above and are reported separately from the page-navigation target. The PR3 installed-package harness reaches Data Info in a 2.557-second median with a schema v2 thin CRB, down from the preceding 4.097-second runtime optimization point; the renderer and navigation timings begin only after that startup phase.

## Measure PR4 gene-major expression and slow pages

PR4 physically stores the exported BPCells matrix in row-major order so a gene query reads the requested rows instead of scanning the cell-major representation. Thin CRB schema v2 and its logical genes-by-cells API remain unchanged. The benchmark alternates the old and new artifacts, uses warm-cache medians, verifies identical dimensions and dimnames, and checks every returned expression result with `all.equal()` before recording a timing.

```{bash pr4-gene-major, eval=FALSE}
Rscript tests/bench/benchmark_bpcells_storage_order.R \
  /path/to/column-major.crb \
  /path/to/gene-major.crb \
  tests/bench/results/million_cell_bpcells_quick_pr4.tsv \
  3
```

| Operation | BPCells column-major | Gene-major | Change |
|---|---:|---:|---:|
| Single gene × 1M cells | 3,982 ms | 194 ms | 95.1% faster |
| RGB, 3 genes × 1M cells | 3,984 ms | 202 ms | 94.9% faster |
| Multi-panel, 9 genes × 1M cells | 3,970 ms | 231 ms | 94.2% faster |
| Mean, 100 genes × 1M cells | 3,748 ms | 55 ms | 98.5% faster |

PR4 also reuses spatial render state and defers inactive Viewer pages. The page benchmark uses the same gene-major CRB for PR3 and PR4 so this comparison isolates runtime code changes rather than crediting the new expression layout twice. These are exploratory single warm-cache observations, not publication medians.

| Page | PR3 first visit | Initial PR4 first visit | Change | Initial PR4 target |
|---|---:|---:|---:|---:|
| Groups | 1,273 ms | 1,272 ms | 0.1% faster | pass `<2 s` |
| Overview | 10,198 ms | 10,375 ms | 1.7% slower | fail `<2 s` |
| Gene Expression | 3,620 ms | 2,135 ms | 41.0% faster | fail `<2 s` |
| Immune Repertoire | 7,504 ms | 6,973 ms | 7.1% faster | fail `<2 s` |
| Trajectory | 3,660 ms | 3,050 ms | 16.7% faster | fail `<3 s` by 50 ms |
| HLA & TCR Motifs | 5,913 ms | 5,714 ms | 3.4% faster | fail `<3 s` |
| Coordinated Views | 4,741 ms | 3,272 ms | 31.0% faster | fail `<2 s` |

The focused follow-up matches cell barcodes once, vectorizes receptor parsing, constructs motif edges without a dense adjacency matrix, draws the complete HLA graph through the shared Canvas renderer, enables WebGPU for the million-point Trajectory projection, and defers its secondary panels until the primary projection flushes.

| Page | PR3 first visit | Focused PR4 first visit | Change | Target |
|---|---:|---:|---:|---:|
| Trajectory | 3,660 ms | 2,434 ms | 33.5% faster | pass `<3 s` |
| HLA & TCR Motifs | 5,913 ms | 2,696 ms | 54.4% faster | pass `<3 s` |

| Page | Focused PR4 repeat | Repeat target |
|---|---:|---:|
| Trajectory | 1,266 ms | fail `<500 ms` |
| HLA & TCR Motifs | 126 ms | pass `<500 ms` |

Prepare the deterministic page fixture beside the same BPCells sidecar, then run either the complete page set or the focused HLA/Trajectory gate:

```{bash pr4-pages, eval=FALSE}
PAGES_CRB="$(dirname "$CRB")/cerebro_mouse_brain_1m_pages.crb"

Rscript tests/bench/prepare_viewer_1m_pages.R \
  . "$CRB" "$PAGES_CRB"

Rscript tests/bench/benchmark_viewer_1m_pages.R \
  . "$PAGES_CRB" \
  /tmp/million-cell-pages.tsv 1

VIEWER_PAGES_ONLY=trajectory,hla \
  Rscript tests/bench/benchmark_viewer_1m_pages.R \
  . "$PAGES_CRB" \
  /tmp/million-cell-hla-trajectory.tsv 1
```

The complete raw PR4 tables are stored in `tests/bench/results/million_cell_bpcells_quick_pr4.tsv`, `tests/bench/results/million_cell_pages_quick_pr3_pr4.tsv`, and `tests/bench/results/million_cell_hla_trajectory_focused_pr4.tsv`.

## Measure the renderer alone

The synthetic harness isolates buffer construction, upload, the first frame,
zoom, and RGB recolouring from Shiny and CRB startup:

```{bash renderer-benchmark, eval=FALSE}
Rscript tests/bench/benchmark_million_cell_renderer.R \
  1000000 15 renderer.csv renderer.png
```

The same run used Apple Metal 3 through WebGPU and produced:

| Metric | Result |
|---|---:|
| GPU buffer size | 15.26 MiB |
| Renderer initialization | 5.9 ms |
| Categorical buffer build | 7.6 ms |
| Categorical upload | 10.7 ms |
| First frame | 30.6 ms |
| Pan/zoom median | 8.6 ms |
| Pan/zoom p95 | 12.2 ms |
| RGB buffer build | 9.0 ms |
| RGB upload | 7.8 ms |
| RGB frame median | 9.5 ms |
| RGB frame p95 | 13.0 ms |

# Compare the CRB representation

The compact BPCells CRB no longer serializes the live matrix handle or repeated
cell names. At runtime the Viewer reopens the adjacent BPCells directory,
validates its cell-name checksum, and restores the omitted metadata and
projection row names.

Measure an old and newly exported CRB after warming both files once. Alternating
the order limits systematic page-cache bias:

```{r crb-load-benchmark}
files <- c(
  before = "/path/to/cerebro_mouse_brain_1m_before.crb",
  after = "/path/to/cerebro_mouse_brain_1m_after.crb"
)
invisible(lapply(files, readRDS))

rows <- list()
for (round in seq_len(5L)) {
  order <- if (round %% 2L) names(files) else rev(names(files))
  for (candidate in order) {
    rows[[length(rows) + 1L]] <- data.frame(
      candidate = candidate,
      round = round,
      read_rds_ms = unname(
        system.time(readRDS(files[[candidate]]))["elapsed"] * 1000
      )
    )
  }
}

raw <- do.call(rbind, rows)
aggregate(raw["read_rds_ms"], raw["candidate"], median)
file.info(files)$size / 1024^2
```

| Metric | Before | After | Change |
|---|---:|---:|---:|
| CRB size | 26.83 MiB | 13.92 MiB | 48.1% smaller |
| Warm `readRDS()` median | 1,535 ms | 113 ms | 13.6x faster |

These numbers describe only the serialized `.crb`. The 3.5 GiB BPCells
expression sidecar is unchanged and remains the authoritative expression and
cell-name store.

# Launch the interactive one-million-cell demo

Point `CEREBRO_1M_DEMO_CRB` at a CRB freshly written by 4.6.0 `saveCerebro()` or an exporter, then run the bundled app. Older CRBs remain compatible but calculate the cell fingerprint once per Viewer session; `convertCerebro()` upgrades one in place, while the copy-preserving example above keeps a baseline artifact for comparison.

```{bash launch-demo, eval=FALSE}
CEREBRO_1M_DEMO_CRB=/path/to/cerebro_mouse_brain_1m.crb \
  Rscript -e 'shiny::runApp("inst", host = "127.0.0.1", port = 7451)'
```

Open `http://127.0.0.1:7451`, then select **10x E18 mouse brain (1M)**. The
dataset starts at 10% sampling with point size 1 and opacity 0.5. Increasing the
sampling control to 100% renders all one million points. When the environment
variable is unset, the bundled application behaves exactly as before and does
not show this dataset.

For a direct launch without the bundled dataset switcher:

```{r launch-direct}
devtools::load_all(quiet = TRUE)
launchCerebro(
  mode = "closed",
  crb_file_to_load = c("10x E18 mouse brain (1M)" = crb),
  percentage_cells_to_show = 10,
  point_size = 1,
  point_opacity = 0.5,
  projections_show_hover_info = TRUE
)
```

# Interpretation limits

- The Viewer values are process medians, not confidence intervals.
- `readRDS()` is a warm operating-system-cache measurement, not cold storage.
- The full Viewer includes R loading, Shiny transport, JavaScript decoding, and
  drawing; the synthetic renderer table measures only the browser renderer.
- WebGPU results depend on the browser, driver, and adapter. The recorded run
  used Headless Chrome 152 and Apple Metal 3.
- The one-million-cell subset is suitable for engineering measurements, not a
  new biological analysis of the source experiment.

Remove the detached baseline checkout when it is no longer needed:

```{bash cleanup, eval=FALSE}
git worktree remove ../CerebroNexus-4.5.1-baseline
```
