---
title: "Loading multiple data sets (multi-crb) with a dataset switcher"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Loading multiple data sets (multi-crb) with a dataset switcher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

cerebroAppLite can load **multiple `.crb` files** in a single app. When more
than one file is supplied to `createShinyApp()`, a **"Select dataset:"**
switcher appears in the sidebar so users can move between data sets without
restarting the app. Each data set keeps its own expression data, metadata, and
module-specific slots — including the immune repertoire — so conditional tabs
(such as *Immune Repertoire*) appear or disappear as you switch.

# Quick start

## Single file (unchanged)

Passing a single path behaves exactly as before — no switcher is shown:

```{r, eval = FALSE}
createShinyApp(
  result_dir = file.path(tempdir(), "cerebro_app_single"),
  cerebro_data = c(demo = system.file(
    "extdata/v1.4/example.crb",
    package = "cerebroAppLite"
  ))
)
```

## Multiple files

Pass a **named vector** of `.crb` paths. The three demo data sets shipped with
the package are genuinely different samples — they differ in cell composition,
so the UMAP and cell-type mix change as you switch — and each carries a
biologically plausible immune repertoire (TCR on T cells, BCR on B cells):

```{r, eval = FALSE}
createShinyApp(
  result_dir = file.path(tempdir(), "cerebro_app_multi"),
  cerebro_data = c(
    "PBMC - Full (T+B)"     = system.file("extdata/v1.4/demo_full_tcr_bcr.crb", package = "cerebroAppLite"),
    "PBMC - Healthy (T/NK)" = system.file("extdata/v1.4/demo_healthy_t.crb",    package = "cerebroAppLite"),
    "PBMC - B-cell rich"    = system.file("extdata/v1.4/demo_bcell_rich.crb",   package = "cerebroAppLite")
  )
)
```

| Data set | Cells | Immune repertoire |
|----------|-------|-------------------|
| PBMC - Full (T+B)     | all cells (T + B + Mono) | TCR **and** BCR |
| PBMC - Healthy (T/NK) | T + Monocytes            | TCR only        |
| PBMC - B-cell rich    | B + a few T cells        | BCR only        |

When the app starts, the sidebar shows a dropdown with these three samples.
Switching changes the whole data set — the UMAP, the cell-type composition, and
the *Immune Repertoire* tab (present whenever the sample carries clonotypes).

The three demo data sets are derived from the public 10x Genomics
`vdj_v1_hs_pbmc3` dataset (see `data-raw/README.md` in the package source for
the full, reproducible build).

# Default file selection

By default the **smallest `.crb` file** (by size on disk) is loaded first. Set
`crb_pick_smallest_file = FALSE` to load the first file in the vector instead:

```{r, eval = FALSE}
createShinyApp(
  result_dir = file.path(tempdir(), "cerebro_app_order"),
  cerebro_data = c(
    a = "file_a.crb",
    b = "file_b.crb"
  ),
  crb_pick_smallest_file = FALSE # loads file_a.crb
)
```

Once a file is chosen from the dropdown, that selection persists until the user
switches again.

# Linking to a specific data set via URL

When the app is deployed, a data set can be selected directly from the URL, so
you can share a link that opens straight into one sample. Both a query string
and a path segment are supported:

```
https://your-host/app/?dataset=TCR
https://your-host/app/TCR
```

The token is the **last path segment** (so it works even when the app is mounted
under a sub-path such as `/app/`), or the value of the `dataset` query
parameter. It is matched against the **names** you gave in `cerebro_data` first,
then against the file basename (with or without the `.crb` extension). If no
match is found, the smallest-file default applies.
